Impact of grids and dynamical cores in CESM2.2 on the meteorology and climate of the Arctic

### 2021-03-12 arcticgrids.tex
more text

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% AGUJournalTemplate.tex: this template file is for articles formatted with LaTeX
%
% This file includes commands and instructions
% given in the order necessary to produce a final output that will
% satisfy AGU requirements, including customized APA reference formatting.
%
% You may copy this file and give it your
% article name, and enter your text.
%
%
% Step 1: Set the \documentclass
%
%

@@

General Circulation Models (GCMs) are powerful tools for understanding the meteorology and climate of the high-latitudes, which are among the most sensitive regions on Earth to global and environmental change. Despite their importance, the numerical treatment of polar regions is handled in vastly-different ways due to the so-called \textit{pole-problem} \cite{W2007JMSJ}. The pole-problem refers to instability arising from the convergence of meridians at the polar points on latitude-longitude grids (e.g., Figure~\ref{fig:uni-grids}a). Depending on the numerical design, methods exist to stabilize the pole problem, and latitude-longitude grids may be advantageous for polar processes as structures can be represented with more degrees of freedom than elsewhere in the computational domain. With the recent trend towards globally uniform unstructured grids, any potential benefits of latitude-longitude grids on polar regions may become a relic of the past. In this study, a spectrum of grids and dynamical cores (hereafter referred to as \textit{dycores}) available in the Community Earth System Model, version 2.2 (CESM; \url{http://www.cesm.ucar.edu/models/cesm2/}), from conventional latitude-longitude grids to unstructured grids with globally uniform grid spacing to unstructured grids with regional refinement over the Arctic, are evaluated to understand their impacts on the simulated characteristics of the Arctic, with a special focus on the Greenland Ice Sheet.

In the 1970's the pole problem was largely defeated through wide-spread adoption of efficient spectral transform methods in GCMs. These methods transform grid point fields into a global, isotropic representation in wave space, where linear operators in the equation set can be solved for exactly. While spectral transform methods were still commonly used into the 21st century, more recently, local numerical methods have become desirable for their ability to run efficiently on massively parallel systems. The pole-problem has thus re-emerged in contemporary climate models using latitude-longitude grids, and some combination of reduced grids and polar filters are necessary to ameliorate this instability \cite{JW2010LNCSE}. Polar filters are akin to a band-aid; they subdue the growth of unstable modes by applying additional damping to the solution over polar regions. One might expect that this additional damping reduces the effective resolution such that the resolved scales are similar across the entire domain, but this seems an unlikely outcome in practice.

@@

The goal of this study is to characterize the representation of high-latitiude regions using the spectral-element and finite-volume dycores in CESM2.2, as these models treat the high-latitudes, e.g., the pole-problem, in very different ways. The manuscript is laid out as follows. Section~\ref{sec:methods} consists of documentation of the grids, dycores and physical parameterizations used in this study. The Arctic refined grids were developed by the author, and this section serves as their official documentation. Section~\ref{sec:methods} also contains a description of the experiments along with the observational datasets and post-processing software used for validating the models. Section~\ref{sec:results} contains the results of the experiments and Section~\ref{sec:conclusions} provides some discussion and conclusions.

@@

\section{Methods}\label{sec:methods}
\subsection{Grids}
\subsection{Dynamical cores}
The atmospheric component of CESM2.2, the Community Atmosphere Model, version 6.3 (CAM; \url{https://github.com/ESCOMP/CAM/wiki}), supports a diverse number of atmospheric dynamical cores. These range from dycores using latitude-longitude grids, the finite-volume \cite<FV;>{L2004MWR} and eulerian spectral transform \cite<EUL;>{CETAL006JC} models, and dycores built on unstructured grids, the spectral-element \cite<SE;>{LetAl2018JAMES} and finite-volume 3 \cite<FV3;>{PL2007JCP} models. The EUL dycore is the oldest dycore in CAM, and is the least supported of all the dycores in the current model. FV3 is the newest dycore in CESM, but it was not fully incorporated into CAM at the time this work commenced, and so is omitted from this study. The authors instead focus on the two most well supported dycores in CESM, SE and FV. The goal of this study is to characterize the representation of polar regions using the SE and FV dycores, as these models treat the high-latitudes, e.g., the pole-problem, in very different ways. 
\subsubsection{Finite-volume model}
The FV dycore is assessed using grids with $\Delta x$ of $1^{\circ}$ and $2^{\circ}$ grid spacing ($\Delta x$ refers to the average equatorial grid spacing). These are compared to the equivalent  $1^{\circ}$ SE grid, referred to as ne30pg3. This grid does not refer to the SE dycore proper, but a variant in which the dry dynamics are solved using the SE method, and tracer advection computed using the conservative semi-Lagrangian advection method \cite <SE-CSLAM;>{LTOUNGK2017MWR}. In SE-CSLAM the physical parameterizations are computed on the finite-volume tracer advection grid. Optionally, one can compute the physics on a finite-volume grid that is $\frac{3}{2} \times \Delta x$ of the tracer-advection grid \cite{HETAL2019JAMES}. A comparison of the lower-resolution physics grid version of ne30pg3, referred to as ne30pg2, over the Arctic is included in this study. All variants of the SE dycore discussed herein are available as part of the CESM2.2 release.
\subsubsection{Spectral-element model}
The SE dycore (not SE-CSLAM) supports regional grid refinement via its variable resolution configuration. Two variable resolution meshes were developed as part of the CESM2.2 release that contains grid refinement over the Arctic. The ARCTIC grid is a $\Delta x=1^{\circ}$ grid with $\Delta x=\frac{1}{4}^{\circ}$ regional refinement over the broader Arctic region. The ARCTICGRIS grid is similar to the ARCTIC grid, but additionally contains a patch covering the big island of Greenland with $\Delta x=\frac{1}{8}^{\circ}$ resolution. These Arctic refined grids are depicted in Figure~\ref{fig:vr-grids}. The surface mass balance of the GrIS is compared across all grids and dycores in this study, and the ARCTCGRIS grid is certainly at an advantage due to its higher resolution over the ice sheet.
\subsection{Physical parameterizations}
\subsection{Experimental desing}
\subsection{Observational datasets}
\subsubsection{ERA5}
\subsubsection{LIVVkit 2.1}
\subsection{TempestExtremes}
\subsection{StormCompositer}

>>>>

% AGUJournalTemplate.tex: this template file is for articles formatted with LaTeX
%
% This file includes commands and instructions
% given in the order necessary to produce a final output that will
% satisfy AGU requirements, including customized APA reference formatting.
%
% You may copy this file and give it your
% article name, and enter your text.
%
%
% Step 1: Set the \documentclass
%
%

@@

General Circulation Models (GCMs) are powerful tools for understanding the meteorology and climate of the high-latitudes, which are among the most sensitive regions on Earth to global and environmental change. Despite their importance, the numerical treatment of polar regions is handled in vastly-different ways due to the so-called \textit{pole-problem} \cite{W2007JMSJ}. The pole-problem refers to instability arising from the convergence of meridians at the polar points on latitude-longitude grids (e.g., Figure~\ref{fig:uni-grids}a). Depending on the numerics, methods exist to stabilize the pole problem, and latitude-longitude grids may be advantageous for polar processes as structures can be represented with more degrees of freedom than elsewhere in the computational domain. With the recent trend towards globally uniform unstructured grids, any potential benefits of latitude-longitude grids on polar regions may become a relic of the past. In this study, a spectrum of grids and dynamical cores (hereafter referred to as \textit{dycores}) available in the Community Earth System Model, version 2.2 (CESM; \url{http://www.cesm.ucar.edu/models/cesm2/}), from conventional latitude-longitude grids to unstructured grids with globally uniform grid spacing to unstructured grids with regional refinement over the Arctic, are evaluated to understand their impacts on the simulated characteristics of the Arctic, with a special focus on the Greenland Ice Sheet.

In the 1970's the pole problem was largely defeated through wide-spread adoption of efficient spectral transform methods in GCMs. These methods transform grid point fields into a global, isotropic representation in wave space, where linear operators in the equation set can be solved for exactly. While spectral transform methods are still commonly used in the 21st century, local numerical methods have become desirable for their ability to run efficiently on massively parallel systems. The pole-problem has thus re-emerged in contemporary climate models that use latitude-longitude grids, and some combination of reduced grids and polar filters are necessary to ameliorate this instability \cite{JW2010LNCSE}. Polar filters are akin to a band-aid; they subdue the growth of unstable modes by applying additional damping to the solution over polar regions. One might expect that this additional damping reduces the effective resolution such that the resolved scales are similar across the entire domain, but this seems an unlikely outcome and is explored further in this study.

@@

The goal of this study is to characterize the representation of high-latitude regions using the spectral-element and finite-volume dycores in CESM2.2, as these models treat the high-latitudes, e.g., the pole-problem, in very different ways. The manuscript is laid out as follows. Section~\ref{sec:methods} consists of documentation of the grids, dycores and physical parameterizations used in this study. The Arctic refined grids were developed by the authors, and this section serves as their official documentation in CESM2.2. Section~\ref{sec:methods} also contains a description of the experiments along with the observational datasets and post-processing software used for evaluating the models. Section~\ref{sec:results} contains the results of the experiments and Section~\ref{sec:conclusions} provides some discussion and conclusions.

@@

\section{Methods}\label{sec:methods}
\subsection{Grids and dynamical cores}

The atmospheric component of CESM2.2, the Community Atmosphere Model, version 6.3 (CAM; \url{https://ncar.github.io/CAM/doc/build/html/index.html}), supports a diverse number of atmospheric dynamical cores. These range from dycores using latitude-longitude grids, the finite-volume \cite<FV;>{L2004MWR} and eulerian spectral transform \cite<EUL;>{CETAL006JC} models, and dycores built on unstructured grids, the spectral-element \cite<SE;>{LetAl2018JAMES} and finite-volume 3 \cite<FV3;>{PL2007JCP} models. The EUL dycore is the oldest dycore in CAM, and the least supported of all the dycores in the current model. FV3 is the newest dycore in CAM, but it was not fully incorporated at the time this work commenced, and so is omitted from this study. The authors instead focus on the two most well supported dycores in CAM, SE and FV.

\subsubsection{Finite-volume model}

The FV dycore is a hydrostatic model that integrates the equations of motion using a finite-volume discretization on a global latitude-longitude grid. The 2D dynamics evolve in floating Lagrangian layers that are periodically mapped to Eulerian reference grid in the vertical. Hyperviscous damping is applied to the divergent modes while Laplacian damping is applied to momentum in the top few layers, referred to as a ``sponge layer." A polar filter is used to avoid computational instability due to the convergence of the meridians, permitting the use of a more practical time-step. It takes the form of a Fourier filter in the zonal direction, with the damping coefficients increasing monotonically poleward.  The FV dycore is run with grids $\Delta x$ of $1^{\circ}$ and $2^{\circ}$ grid spacing, referred to as $f09$ and $f19$, respectively (Figure~\ref{fig:uni-grids}a,b).

\subsubsection{Spectral-element model}

The SE dycore is a hydrostatic model that integrates the equations of motion using a high-order continuous Galerkin method. The computational domain is defined by a cubed-sphere grid that is tiled with quasi-equal area quadralateral elements (e.g., Figure~\ref{fig:vr-grids}). Each element contains a fourth order basis set in the two horizontal directions, and the solution defined at the roots of the basis functions, the Gauss-Lobatto-Lagendre (GLL) quadrature points. This results in 16 GLL point values within each element, with 12 of the points lying on the element boundary. This compact support allows for efficient parallelization up to one element for task. Communication between elements happens via the direct stiffness summation, which applies a numerical flux to the element boundaries to reconcile overlapping nodal point values, and reulting in a continuous global basis set.

As with the FV dycore, the dyanmics evolve in floating Lagrangian layers that are subsequently mapped to a Eulerian reference grid. The 2D dynamics have no implicit dissipation and so hyperviscosity operators are applied to all prognostic variables to remove spurious numerical errors. Laplacian damping is applied in the sponge layer, similar to the FV dycore.

The SE dycore supports regional grid refinement via its variable-resolution configuration. Two variable resolution meshes were developed as part of the CESM2.2 release that contains grid refinement over the Arctic. The software package SQuadgen (\url{https://github.com/ClimateGlobalChange/squadgen}) was used to generate these meshes. The $ARCTIC$ grid is a $\Delta x=1^{\circ}$ grid with $\Delta x=\frac{1}{4}^{\circ}$ regional refinement over the broader Arctic region. The $ARCTICGRIS$ grid is similar to the $ARCTIC$ grid, but additionally contains a patch covering the big island of Greenland with $\Delta x=\frac{1}{8}^{\circ}$ resolution. Both grids are depicted in Figure~\ref{fig:vr-grids}.

For spectral-element grids with quasi-uniform grid spacing, a variant in which tracer advection is computed using the Conservative Semi-Lagrangian Mulit-tracer tranport scheme is used instead \cite[CSLAM]{LTOUNGK2017MWR}. CSLAM has improved tracer property preservation and accelerated multi-tracer transport. It uses a seperate grid from the spectral-element dynamics, through dividing each element into $3\times3$ quasi-equal area control volumes. The physical parameterizations are computed from the state on the CSLAM grid, which has clear advantages over the default SE dycore in which the physics are evaluated at the GLL nodal points \cite{HL2018MWR}.

The $1^{\circ}$ CAM-SE-CSLAM grid is referred to as $ne30pg3$ (Figure~\ref{fig:uni-grids}c). The $ne$ indicates the number of elements along an edge of a cubed-sphere face, and the $pg$ denotes the number of control volumes along the edge of an element used for computing the physical parameterizations. Optionally, one can compute the physics on a finite-volume grid that is $\frac{3}{2} \times \Delta x$ of the tracer-advection grid, found through dividing the elmenet into $2\times2$ control volumes \cite[(Figure~\ref{fig:uni-grids}d)]{HETAL2019JAMES}. A comparison of this lower-resolution physics grid $ne30pg2$ over the Arctic is included in this study. All variants of the SE dycore discussed herein are available as part of the CESM2.2 release.

\subsection{Physical parameterizations}

The CAM6 physical parameterization package (hereafter referred to as the \textit{physics}; \url{https://ncar.github.io/CAM/doc/build/html/index.html}) is used for all simulations in this study. CAM6 physics is most noteably different from it's predecessors through the incorporation of high-order turbulence closure, Cloud Layers Unified by Binormals \cite[CLUBB]{GETAL2002JAS,BOG2013JCLIM}, which jointly acts as a PBL, shallow convection and cloud macrophysics scheme. CLUBB is coupled with the MG2 microphysics scheme \cite{MG2}, with prognostic precipitation and classical nucleation theory in representing cloud ice for improved cloud-aerosol interactions. Deep convection is parameterized using a convective quasi-equilibrium mass flux scheme \cite{ZM1995AO,NRJ2008JC} inlcuding convective momentum transport \cite{RSG2010JAS}. PBL form drag is modeled after \cite{BBW2004QJRMS} and orographic gravity wave drag is represented with an anisotropic method driven by the orientation of topographic ridges at the sub-grid scale.

\subsection{Experimental design}

All grids and dycores are run using an identical transient 1979-1998 AMIP-style configuration, with prescribed monthly SST/sea-ice after \cite{CESMSST}. This configuration refers to the $FHIST \ compset$ and runs out of the box in CESM2.2. The Community Land Model (CLM5; \url{https://escomp.github.io/ctsm-docs/versions/release-clm5.0/html/users_guide/index.html}) calculates the surface energy balance at each land tile within a grid cell, which is used to compute snow and bare ice melting to inform the surface mass balance (SMB) of a glacier unit \cite{VKETAL2020JGR}. Ice accumulation is modeled as a capping flux, or snow in excess of the assumed 10 m snow cap, and refreezing of liquid within the snowpack additionally acts as a source of mass in the SMB calculation. The SMB of the GrIS was spun-up for the variable-resolution grids by forcing CLM5 offline, cycling over 20 years of a fully coupled $ARCTIC$ run for about 500 years. The uniform resolution grids are all initialized with an SMB from an existing $f09$ spun-up initial condition.

\subsection{Observational datasets}
\subsubsection{ERA5}
\subsubsection{LIVVkit 2.1}
\subsection{TempestExtremes}
\subsection{StormCompositer}
